---
title: "Example Analysis: Ebola"
date: "today"
format: html
---

# Preamble

This script will loosely follow the template document: it is more meant for you to get an idea of one way to approach the data processing and visualization! This analysis will primarily focus on using tidyverse tools (i.e. `dplyr` and `ggplot`).

------------------------------------------------------------------------

For this dataset, the **main skills** we wanted you to practice were:

1.  Working with dates: you can get a more fine-scale understanding of the data by parsing out the `Date` column into year, month, and day.
2.  Pivoting: notice that the counts are separated into separate columns (cases and deaths) -- it'll be easier to work with them if they were consolidated.

The **research questions** we will address are:

1.  Easier: How many cases and deaths in total were recorded by each country from 2014-2016?
2.  Harder: By country, how did the average number of cases and death change each year?

# 1. Load data and libraries

```{r}
library(tidyverse)

ebola <- read.csv("scripts/04_projects/project-day-2_files/datasets/ebola.csv", header = TRUE, sep = ",")
```

# 2. Format the Data

## Pre-processing

First, let's just focus on our 3 countries of interest and use `filter()` to trim down our dataset. We'll also rename some of the columns to more manageable names, since they're pretty unwieldly in the raw dataset.

```{r}
ebola <- ebola |>
  filter(Country %in% c("Guinea", "Liberia", "Sierra Leone")) |>
  rename(Cases = Cumulative.no..of.confirmed..probable.and.suspected.cases,
                Deaths = Cumulative.no..of.confirmed..probable.and.suspected.deaths)
```

## Skill: working with dates

Let's parse out the `Date` into separate `Year`, `Month`, and `Day` columns using `mutate()` with the following steps:

1.  Tell R that `Date` contains date objects, using `as.Date().`
    a.  Because of the way our data is formatted, they will automatically take on the `%Y-%m-%d` format.
2.  Create `Year`, `Month`, and `Day` columns by extracting the relevant part of the date object using `format()`.
3.  Simultaneously, we will also convert the parsed data into numeric data using `as.numeric()`, because `format()` will extract that information as a character object.

Remember that there are multiple ways to approach this![^1]

[^1]: <https://stackoverflow.com/questions/4078502/split-date-data-m-d-y-into-3-separate-columns>

```{r}
ebola <- mutate(ebola,
                Date = as.Date(Date), # format the dates
                Year = as.numeric(format(Date, "%Y")), # extract different parts
                Month = as.numeric(format(Date, "%m")),
                day = as.numeric(format(Date, "%d")))
```

## Skill: pivoting

This data doesn't quite follow tidy data principles: we have two types of numeric data, `Cases` and `Death`, spread across the column headers. We'll consolidate them with `pivot_longer()`, converting the column names to `Type` and numbers in the cells into `Count`.

```{r}
ebola <- pivot_longer(ebola,
                      cols = c("Cases", "Deaths"), # select columns to be pivoted
                      names_to = "Type",
                      values_to = "Count")
```

# 3. Visualize the Data

## Easier

::: callout-note
How many cases and deaths in total were recorded by each country from 2014-2016?
:::

We don't need to do anything further to the data: since we want to show totals, a bar graph for each region should suffice.

```{r}
ggplot(ebola, aes(x = Country, y = Count, fill = Type)) +
  geom_col(position="dodge") +
  labs(title = "Total Ebola Cases and Deaths from 2013-2016")
```

## Harder

::: callout-note
By country, how did the average number of cases and death change each year?
:::

We can use `summarise()` to generate some summary statistics. Since we want to look at trends over time, a line plot would be useful.

```{r}
# calculate stats by country, type, year
ebola_stats <- ebola |>
  group_by(Country, Type, Year) |>
  summarise(Average = mean(Count))

# plot it
ggplot(ebola_stats, 
       aes(y = Average, x = Year, color = Country)) +
  geom_line() + geom_point() +
  facet_wrap(~Type) +
  labs(title = "Average Ebola Cases and Deaths from 2013-2016")
```

This looks a little ugly with the decimal points: we can clean it up by converting `Year` to a factor. However, this also confuses R in `geom_line()`, and it can't figure out which points to connect together: so, to clarify, we'll also specify the `group` variable in `aes()`.

```{r}
ggplot(ebola_stats, 
       aes(y = Average, x = as.factor(Year), color = Country,
       group = Country)) +
  geom_line() + geom_point() +
  facet_wrap(~Type) +
  labs(title = "Average Ebola Cases and Deaths from 2013-2016",
       x = "Year")
```
