---
title: "Analysis Template"
format: html
---

## Dataset: [Fill in your dataset here!]


### 1. Create a directory for your analysis



### 2. Load in the data

Don't forget to load in any R libraries that you may need!


```{r load-data}
# YOUR CODE GOES HERE

```



### 3. Data exploration

Some things you may want to look at:
1. How many rows? How many columns?
2. Is each sample 1 row or 1 column?
3. Does the dataset have any missing values? If so, how do you want to deal with them?

```{r explore-data}
# YOUR CODE GOES HERE

```

### 4. Research Question

*** Fill in your research question here! ***



### 5. Formatting the Data

Depending on your research question, what information is important? Is the data in the format you need it to be? E.g. Date formats come in many different forms, and maybe you only care about the year and not month/day!

```{r format-data}
# YOUR CODE GOES HERE

```

### 6. Visualize the Data

Now that your data is in a useful format, try creating some visualizations that can help you answer your research question. What type of plot would be most useful here? Are your different variables numerical or categorical?

```{r visualize-data}
# YOUR CODE GOES HERE

```

#### Conclusions:

What did you discover?



### 7. Push your code!

Using the command line or the GitHub interface, push your code to your repository! Add a commit message that will be helpful to future you. And remember that GitHub is for storing code, not data.

